Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-14 surfaces-write-up/surface-write-up.tex
add Shou-Wu to acknowledgements

\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
%\captionsetup[table]{belowskip = 4pt}

@@

\begin{lem}
Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$.  Then $R_D$ is generated in degree at most $\sum_{i=1}^n \ell_i a_i$.
% and relations in degree at most  .\todo{insert numbers}
\end{lem}
\begin{proof}
Let 
\begin{align*}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}.
\end{align*}
We note that $\Sigma$ has extremal rays given by the lattice points $e_i$ corresponding to
\[
	\epsilon_i = \prod_{j\ne i} (f_j)^{-\alpha_j \ell_i a_i} {f_i}^{\ell_i \sum_{j\ne i} \alpha_j a_j}
\]
for each $i\in \{1, \ldots n\}$.  Therefore, applying proposition \ref{prop:cone-generation}, we see $R_D$ is generated in degree less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

@@

\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_
k)$. Then define 
\begin{equation}\label{eqn:define-T=(D)}
	T_=(D) = \left\{i \in \{1, \ldots, m\}: \bida_i \sum_{k=1}^m \bidb_k 
\alpha_k = \bidb_i \sum_{k=1}^m \bida_k \alpha_k \right\},
\end{equation}

\begin{equation}\label{eqn:define-T+(D)}
	T_+(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} 
\bida_i \sum_{k=1}^m \alpha_k \bidb_k > \bidb_i \sum_{k=1}^m \alpha_k \bida_k 
\right\},
\end{equation}

\noindent
and
\begin{equation}\label{eqn:define-T-(D)}
	T_-(D) = \left\{ j \in \{1, \ldots, m\}: \bida_j \sum_{k=1}^m \alpha_
k \bidb_k < \bidb_j \sum_{k=1}^m \alpha_k \bida_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
$R_D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i) \ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{lem}

@@

For $i \in T_=(D)$
define
\[
	\epsilon_i := (\prod_{k \ne i} f_k^{-\alpha_k \ell_i \gcd(\bida_i, \bidb_i)})
	(f_i^{\ell_i \frac{\gcd(\bida_i, \bidb_i)}{\bida_i + \bidb_i}\sum_{k \ne i}
	\alpha_k (\bida_k + \bidb_k)}) \in H^0(\ell_i \gcd(\bida_i, \bidb_i) D).
\]

For $i \in T_+(D)$ and $j \in T_-(D)$, define
\[
	\epsilon_{i, j} := (\prod_{k = 1}^m f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i,j}(\bida_i \bidb_j - 
\bida_j \bidb_i)D)
\]

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \ref{prop:cone-generation}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sigma = \sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

\todo{factor $\phi$ through $\psi$ and $\chi$.}
\begin{lem}
$\ker(\psi)$ is generated in degree at most \todo{insert number}.
\end{lem}

@@

\noindent
for some $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, $\bida_j + \sum_{i = 1}
^n \bida_i c_i = 0$, $\bidb_j + \sum_{i=1}^n \bidb_i c_i = 0$, and $\beta_j \in
k[y_1, \ldots, y_4]$ satisfies $\psi(\beta_j)\in \mathscr{O}(\bida_j,
\bidb_j)$.

@@

Note that we can view the relations from Equation
\ref{eqn:relations-psi} as an element of $S$ given
by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  Let $S'\subseteq (\mathbb{R}_{\ge 0})^{n+1}$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_j + \sum_{i = 1}^m \bida_i c_i = 0$, and $\bidb_j + \sum_{i=1}^n \bidb_i c_i
= 0$.  A careful computation deduces that $\gamma (d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots, c_n)$ is of the form 
\[
	E + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k},
\] 
where $E$ is $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_i \mbox{ if } i\in S\} \cup \{e_{i,j}: j\in S_- \mbox{ if } i \in S_+\} \cup \{e_{j,i}: j\in S_+ \mbox{ if } i \in S_-\}.
\]

Let $M$ be the set of monomials of $\beta_j$.  Then for $\mu\in M$, the corresponding lattice point is of the form
\[
	E_\mu + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k}
\]
where $E_\mu$ is a $\mathbb{R}_{\ge 0}$-combination of elements in
\begin{align*}
			& \{e_j : j\in T_= \cap \{1, 2, 3, 4\}\} \\
	\cup \; 	& \{e_{j, k}: j\in T_= \cap \{1, 2, 3, 4\}, k\in T_-\}\\
	\cup \; 	& \{e_{k,j}: k\in T_+, j\in T_- \cap \{1, 2, 3, 4\}\}.
\end{align*}

Let $F\in R_D$ correspond to $E + \sum_{k\in T_=} (r_k) e_k + \sum_{h\in T_+ \atop k\in T_-} (r_{h,k}) e_{h,k}$ where $(r)$ signifies the fractional part of $r$.  Similarly, let $F_\mu \in R_D$ correspond to $E_\mu + \sum_{h\in T_+ \atop k\in T_-} (r_{h,k}) e_{h,k}$.  Then $F - \sum_{\mu\in M} F_\mu = 0$ gives us a relation in degree at most
\[
	\sigma + \deg(F),
\]
which generates $(z_i - \beta_i) \prod_{i=1}^n y_i^{c_i}$
We can bound $\deg(F)$ by
\[
	\deg(F) \le \left\{ \begin{split}  \ell_i \gcd(\bida_i, \bidb_i) \mbox{ if } i \in S_= 
			\\ \max_{j\in S_-}(\ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)) \mbox{ if } i\in S_+ \\
			\\ \max_{j\in S_+}(\ell_{j, i} (\bida_j \bidb_
i - \bida_i \bidb_j)) \mbox{ if } i \in S_- \end{split}  \right.
\]
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max( \max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{i\in S_+ \atop j\in S_-}(\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))).
\]

>>>>

\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
\DeclareMathOperator{\Te}{T_=}
\DeclareMathOperator{\Tp}{T_+}
\DeclareMathOperator{\Tm}{T_-}
%\captionsetup[table]{belowskip = 4pt}

@@

\begin{lem} \label{P-m-generators}
Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$.  Then $R_D$ is generated in degree at most $\sum_{i=1}^n \ell_i a_i$.
% and relations in degree at most  .\todo{insert numbers}
\end{lem}
\begin{proof}
Let 
\begin{align*}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}.
\end{align*}
We note that $\Sigma$ has extremal rays given by the lattice points $e_i$ corresponding to
\[
	\epsilon_i = \prod_{j\ne i} (f_j)^{-\alpha_j \ell_i a_i} {f_i}^{\ell_i \sum_{j\ne i} \alpha_j a_j}
\]
for each $i\in \{1, \ldots n\}$.  Therefore, applying proposition \ref{prop:cone-generation}, we see $R_D$ is generated in degree less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

\todo{Discussion of factoring $\phi$ through $\psi$ and $\chi$}

\begin{lem}\label{P-m-relations-psi}
$\ker(\psi)$ is generated in degree at most \todo{insert number}.
\end{lem}
\begin{proof}
We first claim that $\psi$ has a relation of the form
\begin{equation}\label{eqn:relations-psi}
	(z_i - \beta_j)\prod_{j=1}^n {z_j}^{c_{i}}
\end{equation}

\noindent
for some $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, $\bida_j + \sum_{i = 1}
^n \bida_i c_i = 0$, $\bidb_j + \sum_{i=1}^n \bidb_i c_i = 0$, and $\beta_j \in
k[z_1, \ldots, y_3]$ satisfies $\psi(\beta_j)\in \mathscr{O}(\bida)$.

To find such a $\beta$, we simply write $\epsilon_j$ as a 
polynomial in $\epsilon_1, \ldots \epsilon _3$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{i =
1}^n \epsilon_i^{r_i}$ to a canonical form. \todo{add reference to
earlier description of canonical form.}
\end{proof}

@@

\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_
k)$. Then define 
\begin{equation}\label{eqn:define-T=(D)}
	\Te(D) = \left\{i \in \{1, \ldots, m\}: \bida_i \sum_{k=1}^m \bidb_k 
\alpha_k = \bidb_i \sum_{k=1}^m \bida_k \alpha_k \right\},
\end{equation}

\begin{equation}\label{eqn:define-T+(D)}
	\Tp(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} 
\bida_i \sum_{k=1}^m \alpha_k \bidb_k > \bidb_i \sum_{k=1}^m \alpha_k \bida_k 
\right\},
\end{equation}

\noindent
and
\begin{equation}\label{eqn:define-T-(D)}
	\Tm(D) = \left\{ j \in \{1, \ldots, m\}: \bida_j \sum_{k=1}^m \alpha_
k \bidb_k < \bidb_j \sum_{k=1}^m \alpha_k \bida_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
$R_D$ is generated in degree at most
\[
	\sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i) \ell_i + \sum_{i\in \Tp(D) \atop j\in \Tm(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{lem}

@@

For $i \in \Te(D)$
define
\[
	\epsilon_i := (\prod_{k \ne i} f_k^{-\alpha_k \ell_i \gcd(\bida_i, \bidb_i)})
	(f_i^{\ell_i \frac{\gcd(\bida_i, \bidb_i)}{\bida_i + \bidb_i}\sum_{k \ne i}
	\alpha_k (\bida_k + \bidb_k)}) \in H^0(\ell_i \gcd(\bida_i, \bidb_i) D).
\]

For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\[
	\epsilon_{i, j} := (\prod_{k = 1}^m f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i,j}(\bida_i \bidb_j - 
\bida_j \bidb_i)D)
\]

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \ref{prop:cone-generation}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in \Tp(D) \atop
	j \in \Tm(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

\todo{factor $\phi$ through $\psi$ and $\chi$.}
\begin{lem}
$\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max( \max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{i\in S_+ \atop j\in S_-}(\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))).
\]
\end{lem}

@@

\noindent
for some $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, $\bida_j + \sum_{i = 1}
^n \bida_i c_i = 0$, $\bidb_j + \sum_{i=1}^n \bidb_i c_i = 0$, and $\beta_j \in
k[z_1, \ldots, z_4]$ satisfies $\psi(\beta_j)\in \mathscr{O}(\bida_j,
\bidb_j)$.

@@

Note that we can view the relations from Equation
\ref{eqn:relations-psi} as an element of $S$ given
by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  Let $S'\subseteq (\mathbb{R}_{\ge 0})^{n+1}$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_j + \sum_{i = 1}^m \bida_i c_i = 0$, and $\bidb_j + \sum_{i=1}^n \bidb_i c_i
= 0$.  A careful computation deduces that $\gamma (d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots, c_n)$ is of the form 
\[
	E + \sum_{k\in \Te} r_k e_k + \sum_{h\in \Tp \atop k\in \Tm} r_{h,k} e_{h,k},
\] 
where $E$ is $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_i \mbox{ if } i\in S\} \cup \{e_{i,j}: j\in S_- \mbox{ if } i \in S_+\} \cup \{e_{j,i}: j\in S_+ \mbox{ if } i \in S_-\}.
\]

Let $M$ be the set of monomials of $\beta_j$.  Then for $\mu\in M$, the corresponding lattice point is of the form
\[
	E_\mu + \sum_{k\in \Te} r_k e_k + \sum_{h\in \Tp \atop k\in \Tm} r_{h,k} e_{h,k}
\]
where $E_\mu$ is a $\mathbb{R}_{\ge 0}$-combination of elements in
\begin{align*}
			& \{e_j : j\in \Te \cap \{1, 2, 3, 4\}\} \\
	\cup \; 	& \{e_{j, k}: j\in \Te \cap \{1, 2, 3, 4\}, k\in \Tm\}\\
	\cup \; 	& \{e_{k,j}: k\in \Tp, j\in \Tm \cap \{1, 2, 3, 4\}\}.
\end{align*}

Let $F\in R_D$ correspond to $E + \sum_{k\in \Te} (r_k) e_k + \sum_{h\in \Tp \atop k\in \Tm} (r_{h,k}) e_{h,k}$ where $(r)$ signifies the fractional part of $r$.  Similarly, let $F_\mu \in R_D$ correspond to $E_\mu + \sum_{h\in \Tp \atop k\in \Tm} (r_{h,k}) e_{h,k}$.  Then $F - \sum_{\mu\in M} F_\mu = 0$ gives us a relation in degree at most
\[
	\sigma + \deg(F),
\]
which generates $(z_i - \beta_i) \prod_{i=1}^n y_i^{c_i}$
We can bound $\deg(F)$ by
\[
	\deg(F) \le \left\{ \begin{split}  \ell_i \gcd(\bida_i, \bidb_i) \mbox{ if } i \in S_= 
			\\ \max_{j\in S_-}(\ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)) \mbox{ if } i\in S_+ \\
			\\ \max_{j\in S_+}(\ell_{j, i} (\bida_j \bidb_
i - \bida_i \bidb_j)) \mbox{ if } i \in S_- \end{split}  \right.
\]
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max( \max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{i\in S_+ \atop j\in S_-}(\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))).
\]
